---
title: "Homework #5: Density Estimation" 
author: "**Nicholas Keeley**"
date: "Due: Thu Oct 21 | 3:25pm"
output: R6030::homework
---

NOTE: I referenced Lecture 14 for the KDE code in this assignment, and consulted classmate Geoff Hansen on the derivation math for problem 1.a.

**DS 6030 | Fall 2021 | University of Virginia**

------------------------------------------------------------------------

```{r config, echo=FALSE}
source(system.file("config/hw_config.R", package="R6030")) # knitr settings
# options(dplyr.summarise.inform = FALSE)  # ignore dplyr message about grouping
```


# Required R packages and Directories

::: {.solution}
```{r packages, message=FALSE, warning=FALSE}
data.dir = 'https://mdporter.github.io/DS6030/data/' # data directory
library(R6030)     # functions for SYS-6018
library(ks)        # functions for KDE
library(tidyverse) # functions for data manipulation   
```
:::


# Problem 1 Geographic Profiling

Geographic profiling, a method developed in criminology, can be used to estimate the [home location (roost) of animals](https://www.sciencedirect.com/science/article/pii/S0022519305004157) based on a collection of sightings. The approach requires an estimate of the distribution the animal will travel from their roost to forage for food. 

A sample of $283$ distances that pipistrelle bats traveled (in meters) from their roost can be found at: 

- **Bat Data**: <https://mdporter.github.io/DS6030/data/geo_profile.csv>


One probability model for the distance these bats will travel is:
\begin{align*}
f(x; \theta) = \frac{x}{\theta} \exp \left( - \frac{x^2}{2 \theta} \right)
\end{align*}
where the parameter $\theta > 0$ controls how far they are willing to travel. 


## a. Derive the MLE for $\theta$ (i.e., show the math). 

::: {.solution}

\begin{align*}
\ln(L(\theta)) = \sum_{i=1}^{n} \ln(f(x_{i}; \theta))
\end{align*}

\begin{align*}
1) \hspace{12pt} = \ln(\frac{x_{i}}{\theta} + \ln \exp \left( - \frac{x_{i}^2}{2 \theta} \right)
\end{align*}

\begin{align*}
2) \hspace{12pt} = \ln(x_{i}) - \ln(\theta) - \left(\frac{x_{i}^2}{2 \theta} \right)
\end{align*}

\begin{align*}
3) \hspace{12pt} \frac{\partial}{\partial \theta} \sum_{i=1}^{n} \ln(x) - \ln(\theta) - \left(\frac{x^2}{2 \theta} \right) = \sum_{i=1}^{n} -\frac{1}{\theta} + \left(\frac{x^2}{2 \theta^2} \right)
\end{align*}

\begin{align*}
4) \hspace{12pt} \theta_{MAX} = \sum_{i=1}^{n} -\frac{1}{\theta} + \left(\frac{x^2}{2 \theta^2} \right) = 0
\end{align*}

\begin{align*}
5) \hspace{12pt} = \frac{1}{2\theta} \sum_{i=1}^{n} x^2 = \frac{n}{\theta}
\end{align*}

\begin{align*}
6) \hspace{12pt} \theta_{MAX} = \frac{1}{2n} \sum_{i=1}^{n} x^2
\end{align*}


:::


## b. What is the MLE of $\theta$ for the bat data? (Use results from a, or use computational methods.) 

::: {.solution}

```{r echo=FALSE, results='markup'}

## Read the data

df=read_csv("https://mdporter.github.io/DS6030/data/geo_profile.csv", col_names = FALSE)
df=tibble(meters=df$X1)

## Create data frame with sum log-likelihoods for a sequence of thetas

x=df$meters
theta.seq=seq(0.1,20,length=30)
ntheta=length(theta.seq)
loglike=numeric(ntheta)

for(i in 1:ntheta){
  pfunction=sum(log((x/theta.seq[i])*exp(-(x^2)/(2*theta.seq[i]))))
  loglike[i]=pfunction
}

loglike=tibble(loglikelihood=loglike,
               theta=theta.seq)

## Print output

plot(loglike$theta, loglike$loglikelihood, xlab="Theta", ylab="Log Likelihood", type="p")


theta_MLE=theta.seq[which.max(loglike$loglikelihood)]
paste("The MLE for theta is: ", theta_MLE)
```


:::


## c. Using the MLE value of $\theta$ from part b, compute the estimated density at a set of evaluation points between 0 and 8 meters. Plot the estimated density.

::: {.solution}

```{r}

## Plot sequence of x values between 0-8 to see density

x.seq=seq(0,8,length=100)

prob=numeric(length(x.seq))
for(i in 1:length(x.seq)){
  pfunction=(x.seq[i]/theta_MLE)*exp(-(x.seq[i]^2)/(2*theta_MLE))
  prob[i]=pfunction
}
result=tibble(x=x.seq,
              probability=prob)
plot(result$x, result$probability, xlab="Meters", ylab="Density")
#hist(df$meters)
```


:::


## d. Estimate the density using KDE. Report the bandwidth you chose and produce a plot of the estimated density. 


::: {.solution}

## Lower case h for univariate
## Upper case H for MV
```{r echo=FALSE, results='markup'}

## Estimate density using KDE

f.kde=kde(df$meters)
plot(f.kde)

## Show chosen h

paste("Bandwith was chosen using HPI method: ", f.kde$h)

```


:::


## e. Which model do you prefer, the parametric or KDE? 

::: {.solution}

I prefer the KDE model because it provides more granularity than the parametric model through providing local density around each point, rather than applying a uniform or rigid parametric function that applies equal density weights around each point.

:::




# Problem 2: Interstate Crash Density

Interstate 64 (I-64) is a major east-west road that passes just south of Charlottesville. Where and when are the most dangerous places/times to be on I-64? The crash data (link below) gives the mile marker and fractional time-of-week for crashes that occurred on I-64 between mile marker 87 and 136 in 2016. The time-of-week data takes a numeric value of *\<dow\>.\<hour/24\>*, where the dow starts at 0 for Sunday (6 for Sat) and the decimal gives the time of day information. Thus `time=0.0417` corresponds to Sun at 1am and `time=6.5` corresponds to Sat at noon. 

- **Crash Data**: <`r file.path(data.dir, 'crashes16.csv')`>


## a. Extract the crashes and make a scatter plot with mile marker on x-axis and time on y-axis. 
::: {.solution}

```{r echo=FALSE}

df2=read_csv("https://mdporter.github.io/DS6030/data//crashes16.csv")
plot(df2$mile, df2$time, xlab="Mile Marker", ylab="Timestamp")
```


:::


## b. Use KDE to estimate the *mile marker* density. 

- Report the bandwidth. 
- Plot the density estimate.

::: {.solution}

```{r}

## Bandwidth

b.kde=kde(df2$mile)
paste("Bandwith was chosen using HPI method: ", b.kde$h)

```


```{r}

## KDE estimate

plot(b.kde, xlab="Mile Marker")

```


:::


## c. Use KDE to estimate the temporal *time-of-week* density. 

- Report the bandwidth. 
- Plot the density estimate. 

:::: {.solution}
```{r}

## Bandwidth

c.kde=kde(df2$time)
paste("Bandwith was chosen using HPI method: ", c.kde$h)

```


```{r}

## KDE estimate

plot(c.kde, xlab="Timestamp")
```
:::


## d. Use KDE to estimate the bivariate mile-time density. 

- Report the bandwidth parameters.
- Plot the bivariate density estimate. 

::: {.solution}

```{r}

## Bandwidth

d.kde=kde(df2)
d.kde$H
paste("Bandwith was chosen using HPI method: ", d.kde$H)
print(d.kde$H)

```
```{r}
## Density plot

plot(d.kde)
```


:::


## e. Based on the estimated density, approximate the most dangerous mile marker and time-of-week. 

::: {.solution}

The most dangerous mile marker/combination on this highway appears to be mile marker 118 at 12:43PM on Monday.  

```{r}

## Milemark with highest probability of crashes (indp. of time)

b.kde$eval.points[which.max(b.kde$estimate)]

## Time with highest probability of crashes (indp. of milemarker)

c.kde$eval.points[which.max(c.kde$estimate)]

```



:::

		